Matroids from families of null hypotheses

### 2013-08-19 presentation/presentation.tex
fixed a couple of typos, added some slides towards one-sided hypotheses

% Created 2013-07-25 Thu 18:08
\documentclass[bigger]{beamer}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usetikzlibrary{decorations,arrows,shapes}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{bm}
\tolerance=1000
\usepackage{amsmath} \usepackage{algorithm} \usepackage{graphicx}
\usepackage{amssymb} \usepackage{mathrsfs} 
\newcommand{\bs}[1]{\bm{#1}}
\newcommand{\ie}{{\em i.e.},~}
\newcommand{\eg}{{\em e.g.},~}
\newcommand{\parentd}{../}

\providecommand{\alert}[1]{\textbf{#1}}
\usetheme{Novartis}
\doNotUseNovartisLogoOnAllSlides
\doNotUseNovartisLogoOnTitlePage

@@

\begin{tabular}{ll}
  \begin{minipage}{.3\textwidth}
    \includegraphics<1>[width=1.2\textwidth]{lattice4.pdf}
    \includegraphics<2>[width=1.2\textwidth]{lattice4accept.pdf}
    \includegraphics<3>[width=1.2\textwidth]{lattice4reject.pdf}
    \includegraphics<4>[width=1.2\textwidth]{lattice4rejacc.pdf}\\
    {\tiny wikipedia:Partition\_of\_a\_Set}
  \end{minipage} &
  \begin{minipage}{.7\textwidth}
  \begin{itemize}
  \item All flats are given by intersections of hyperpanes
    $\rightarrow$ knowing all hyperplanes remaining flats are given by
    intersection of index sets
  \item Hyperplanes can by found by algorithms that enumerate the
    circuits of the dual matroid \cite{boros2003algorithms}
  \item The partial order of flats defined by inclusion defines a
    geometric lattice
  \item<2-> Testing only a subset of elementary hypotheses corresponds
    to intervals
  \item<3-> Rejection of elementary hypotheses removes intervals
  \end{itemize}

@@

\begin{frame}{Matroid procedures for pairwise mean comparisons}
\begin{tabular}{ll}
  \begin{minipage}{.3\textwidth}
    \includegraphics[width=1.1\textwidth]{setpartitions.pdf} \\
     {\tiny wikipedia:Bell\_number}
  \end{minipage} &
  \begin{minipage}{.7\textwidth}
  \begin{itemize}
  \item Two-sided comparisons of $m$ means
  \item The cycle matroid of the graph representation
    provides equivalent dependence structure
  \item The flats of a (simple) complete graph are
    isomporphic to the partitions of a set of size $m$
  \item Efficient algorithms to generate all exhaustive sets by
    enumeration of set partitions \cite{er1988fast,knuth2005art,kokosinski2006new} 
  \item The number of flats are $m$-th Bellnumber $-1$ \ie 4, 14,
    51, 202, 876, 4139, ... 
  \end{itemize}
  \end{minipage}
\end{tabular}

@@

\textbf{Old wine in new bottles ((c) L. Hothorn)}
\begin{itemize}
\item \cite{shaffer1986modified} represents pairwise comparisons
  by set partitions 
\item \cite{bergmann1988improvements} define the concepts of an
  exhaustive sets
\item \cite{bernhard1991computergestuetzte} shows properties of
  exhaustive sets that are basically {\em closure axioms} of a matroid
  and provides algorithms to check each intersection hypotheses if
  flat using rank function.
\item \cite{westfall1997multiple} checks each intersection hypotheses
  whether it is flat using dependence arguments.
\item \cite{weichert2000robuste} suggest an MTP based on clique
  decomposition of the graph
\item \cite{westfall2007multiple} provide a branch and bound algorithm
  to generate all flats. They note that for pairwise comparisons all
  exhaustive hypotheses are {\em partition hypotheses} 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Matroid representation}
  \begin{itemize}
  \item proofs are shorter (more elegant),
  \item efficient algorithms exist to find the flats of matroids,
  \item {\em Scumm principle}\cite{welsh2010matroid} everything
    important happens at the top (hyperplanes)
  \item we may define procedures using independent sets,
  \item \ldots{}
  \end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Matroid conditions}
\end{frame}

\begin{frame}
\frametitle{Flats <-> Weakliy exhaustive sets}
\end{frame}

\begin{frame}
\frametitle{Flats <-> Strictly exhaustive sets}
\end{frame}
\section{Summary \& Outlook}

\begin{frame}
\frametitle{Summary}
\end{frame}

\begin{frame}
\frametitle{Outlook}

@@

>>>>

% Created 2013-07-25 Thu 18:08
\documentclass[bigger]{beamer}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usetikzlibrary{decorations,arrows,shapes}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{bm}
\usepackage{ulem}
\tolerance=1000
\usepackage{amsmath} \usepackage{algorithm} \usepackage{graphicx}
\usepackage{amssymb} \usepackage{mathrsfs} 
\newcommand{\bs}[1]{\bm{#1}}
\newcommand{\ie}{{\em i.e.},~}
\newcommand{\eg}{{\em e.g.},~}
\newcommand{\parentd}{../}
\newcommand{\R}{\mathbb{R}}
\newcommand{\indf}[1]{\mathbf{1}_{\left\{#1\right\}}}
\providecommand{\alert}[1]{\textbf{#1}}
\usetheme{Novartis}
\doNotUseNovartisLogoOnAllSlides
\doNotUseNovartisLogoOnTitlePage

@@

\begin{tabular}{ll}
  \begin{minipage}{.3\textwidth}
    \includegraphics<1>[width=1.2\textwidth]{lattice4.pdf}
    \includegraphics<2>[width=1.2\textwidth]{lattice4accept.pdf}
    \includegraphics<3>[width=1.2\textwidth]{lattice4reject.pdf}
    \includegraphics<4>[width=1.2\textwidth]{lattice4rejacc.pdf}\\
    {\tiny wikipedia:Partition\_of\_a\_Set}
  \end{minipage} &
  \begin{minipage}{.7\textwidth}
  \begin{itemize}
  \item All flats are given by intersections of hyperplanes
    $\rightarrow$ knowing all hyperplanes remaining flats are given by
    intersection of index sets
  \item Hyperplanes can by found by algorithms that enumerate the
    circuits of the dual matroid \cite{boros2003algorithms}
  \item The partial order of flats defined by inclusion defines a
    geometric lattice
  \item<2-> Testing only a subset of elementary hypotheses corresponds
    to intervals
  \item<3-> Rejection of elementary hypotheses removes intervals
  \end{itemize}

@@

\begin{frame}{Matroid procedures for pairwise mean comparisons}
\begin{tabular}{ll}
  \begin{minipage}{.3\textwidth}
    \includegraphics[width=1.1\textwidth]{setpartitions.pdf} \\
     {\tiny wikipedia:Bell\_number}
  \end{minipage} &
  \begin{minipage}{.7\textwidth}
  \begin{itemize}
  \item Two-sided comparisons of $m$ means
  \item The cycle matroid of the graph representation
    provides equivalent dependence structure
  \item The flats of a (simple) complete graph are
    isomporphic to the partitions of a set of size $m$
  \item Efficient algorithms to generate all exhaustive sets by
    enumeration of set partitions \cite{er1988fast,knuth2005art,kokosinski2006new} 
  \item The number of flats are $m$-th Bell-Number $-1$ \ie 4, 14,
    51, 202, 876, 4139, ... 
  \end{itemize}
  \end{minipage}
\end{tabular}

@@

\textbf{Old wine in new bottles ((c) L. Hothorn)}
\begin{itemize}
\item \cite{shaffer1986modified} represents pairwise comparisons
  by set partitions 
\item \cite{bergmann1988improvements} define the concepts of an
  exhaustive sets
\item \cite{bernhard1991computergestuetzte} shows properties of
  exhaustive sets that are basically {\em closure axioms} of a matroid
  and provides algorithms to check each intersection hypotheses if is
  a flat using rank function.
\item \cite{westfall1997multiple} generates flats using dependence
  arguments in vector spaces.
\item \cite{weichert2000robuste} suggest an MTP based on clique
  decomposition of the graph, cliques in the graph are flats
\item \cite{westfall2007multiple} provide a branch and bound algorithm
  to generate all flats. They note that for pairwise comparisons all
  exhaustive hypotheses are {\em partition hypotheses} 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Matroid representation}
  \begin{itemize}
  \item proofs are shorter (more elegant),
  \item efficient algorithms exist to find the flats of matroids,
  \item {\em Scum principle}\cite{welsh2010matroid} everything
    important happens at the top (hyperplanes)
  \item we may define procedures using independent sets,
  \item \ldots{}
  \end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{General Results}
\begin{itemize}
\item We give sufficient (and probably necessary) conditions for a
  general family of hypotheses $\mathscr{H}$ to define a matroid
\item We give sufficient (and probably necessary) conditions on
  $\mathscr{H}$ for $\mathscr{L}$ to be the set of strictly exhaustive
  sets
\item Systems with one sided hypotheses, and parallel hypotheses (\eg
  combined non-inferiority/superiority) do not give matroids with
  $\mathscr{L}$ all strictly exhaustive 
\end{itemize}

@@

\begin{frame}
  \frametitle{One sided hypotheses}
  \begin{block}{One-sided linear contrast hypotheses}
    A family of {\em one sided linear contrast hypotheses} is defined
    by a contrast matrix $\bs{A} \in \mathbb{R}^{m\times n}$ and
    vector $\bs{b} \in \mathbb{R}^n$ that define a  family of null
    hypotheses of the form $\mathscr{H} = \{H_i: \bs{A_i}^t\bs{\mu}
    \leq b_i,i \in \{1,...,n\}\}$, where $\bs{A}_i$ denotes the $i$-th
    column of $\bs{A}$ and $b_i$ the $i$-th componenth of $\bs{b}$.
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Hyperplane arrangement}
  \begin{itemize}
  \item Each hypothesis $H_i$ corresponds to a hyperplane $h_i =
    \{\bs{x} \in \mathbb{R}^m: \bs{A}_i^t\bs{x} = b_i\}$
  \item Each hyperplane $h_i$ divides $\mathbb{R}^m$ into two regions
    \begin{itemize}
    \item $h_i^+ = \{\bs{x} \in \mathbb{R}^m: \bs{A}_i^t\bs{x} <
      b_i\}$ 
    \item $h_i^- = \{\bs{x} \in \mathbb{R}^m: \bs{A}_i^t\bs{x}
      > b_i\}$
    \end{itemize}
  \item Defines a sign pattern 
    \begin{equation}
      \label{eq:sign}
      \delta_i(\bs{x}) = \left\{
        \begin{array}{ll}
          + & \text{if } \bs{x} \in h_i^+\\
          0 & \text{if } \bs{x} \in h_i \\
          - & \text{if } \bs{x} \in h_i^-\\
        \end{array}\right.
    \end{equation}
  \item $\delta_i$ is $-$ if $H_i$ false at $x$, or $0$ or $+$
    otherwise. $(\delta_1,...,\delta_n)$ tells us to which combination
    of true and false null hypotheses some $\bs{x}$ belongs.

  \item Set of all sign patterns defined by $\mathscr{H}$:
    \begin{equation}
      \label{eq:faces}
      \mathcal{F}(\mathscr{H}) = \{\bs{\delta}(\bs{x}): \bs{x} \in \R^m\}
    \end{equation}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Example}

\begin{tabular}{ll}
  \begin{minipage}{.4\textwidth}
  \includegraphics[height=1\textwidth]{spheresfukuda.eps}
  \end{minipage} &
  \begin{minipage}{.6\textwidth}
    \begin{itemize}
    \item 4 contrast hypotheses about 3 means
    \item Cut the hyperplanes with the sphere in $\R^3$ (makes
      visualisation easier)
    \item Cells of the sphere arrangment correspond to sign patterns
      of $\mathscr{H}$
    \item The abstraction of such an arrangement is an \alert{Oriented Matroid}
    \end{itemize}
  \end{minipage}
\end{tabular}

\begin{frame}
  \frametitle{Stepwise multiple testing procedure for one-sided linear
    contrast hypotheses}

  Consider a family of general contrast hypotheses $\mathscr{H} =
  \{H_i: \bs{A_i}^t\bs{\mu} \leq b_i,i \in \{1,...,n\}\}$ and let
  $\mathscr{F}(\mathscr{H}) = \{ \bs{\delta}(\bs{x}): \bs{x} \in
  \R^m\}$ denote the corresponding set of sign patterns. Consider a
  set of decision rules $\varphi_F: \R^n \rightarrow \{0,1\}$ for
  which hold $P(\varphi_F = 1 | H_i: F_i \geq 0) \leq \alpha$. Then a
  corresponding decision rule for each elementary hypotheses $H_i$ is
  given by 
  \begin{equation}
    \label{eq:signtest}
    \psi_i = \min_{F: F_i \geq 0} \varphi_F,
  \end{equation}
  which controls the FWER at level $\alpha$ in the strong sense.

\begin{frame}
  \frametitle{Computation of $\mathscr{F}$}
  The question remains how to compute $\mathcal{F}$. For this observe
that the $h_i$ define a hyperplane configuration and $\mathcal{F}$ is
just the face lattice of the corresponding oriented
matroid. Consequently, $\mathcal{F}$ can be found using algorithms
for arrangement construction from polyhedral programming
\cite{Avis-KF-92,Avis-KF-96,Ferrez-KF-Liebling-01} which are
polynomial in time and input size and compact in space!

\begin{frame}
  \frametitle{Example}
  3 ordered means 
\end{frame}

\section{Summary \& Outlook}

\begin{frame}
\frametitle{Summary}
\begin{itemize}
\item Matroids provide usefull abstraction for systems of algebraically dependent hypotheses
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Outlook}
\begin{itemize}
\item MCPmod on \sout{steroids} matroids
\end{itemize}
\end{frame}
